---
title: "Project 3"
author: "Yvette Callender"
date: '"` r "Sys.Date()`"'
output: 
    html_document:
        toc: TRUE
        toc_depth: 2
params: 
    education_level: "no_high_school"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following libraries were used in this project.

```{r}
library(tidyverse)
```

# Introduction

In this project, data related to diabetes health was analyzed using a variety of models.  

## Data 
A Diabetes Health Indicators Dataset , `diabetes_binary_health_indicators_BRFSS2015.csv` was used for this project.  This dataset can be found online [on Kaggle.](https://www.kaggle.com/code/jerryodegua/eda-prediction-of-diabetes-data)

## Variables
The dataset contains 21 variables than can potentially be used in the analysis.  A subset was chosen and they are described below.

The variable `diabetes_binary` is the response variable for this analysis. It is either a 0 (corresponding to a lack of either prediabetes or diabetes) or a 1 (corresponding to the presence of either prediabetes or diabetes).

The variable `HighBP` is a binary variable with 1 corresponding to high blood pressure and 0 corresponding to normal blood pressure.

The variable `HighChol` is a binary variable with 1 corresponding to high cholesterol levels and 0 corresponding to normal cholesterol levels.

The variable `BMI` (Body Mass Index) is a numeric variable.  As part of the analysis, it was converted to a categorical variable with values corresponding to underweight, normal, overweight and obese.

The variable `HeartDiseaseorAttack` is a binary variable where 1 indicates a history of heart disease or attack and a 0 indicates a lack thereof.

The variable `GenHlth` is a categorical variable describing the the reported level of general health as described below:

|Level|Description|
|:---|:---------:|
|1|excellent|
|2|very good|
|3|good|
|4|fair|
|5|poor|

The variable `PhysHlth` corresponds to the number of days out of the last 30 that the patient reports physical injury or illness. 

The variable `DiffWalk` is binary with 1 corresponding to reported difficulty walking or climbing stairs and 0 corresponding to no reported 

The variable `Age` corresponds to age in years. As part of this analysis it was transformed into a categorical variable corresponding to 10-year increments. 

The variable `Education` is a categorical variable with the following levels. 

|Education Level|Description|
|:---|:---------:|
|1|Never attended school or only kindergarten|
|2|Grades 1 through 8 (Elementary)|
|3|Grades 9-11 (Some high school)|
|4|Grade 12 or GED (High school graduate)|
|5|College 1 year to 3 years (Some college or technical school)|
|6|College 4 years or more (College graduate)|

For this analysis, data was divided into subsets based on the value of `Education`.  There are be five subsets: one for levels one and two combined, and one each for levels three, four, five, and six.

## Purpose of EDA and Modeling and End Results

The relationship between the presence of diabetes/prediabetes (as indicated by `diabetes_binary`) and the predictor variables `()`HighBP`, `HighChol`, `BMI` , `HeartDiseaseorAttack`, `GenHlth`,  `PhysHlth`, `DiffWalk` were explored at several levels of eduction.  After Exploratory Data Analysis, the relationships were modeled using a variety or approaches (logistic regression, LASSO logistic regression, classification tree, random forest, *other 1, and other 2*) and a best model was chosen for each level of education. 

#Data
#Summarizations
(group means sds, contingency tables six graphs)